### Activity 4: Exercise Solutions ###

Solutions for Exercises in Activity 5
  
#### Exercise 4.1: A simple linear regression model ####
  
1) Look at the Old Faithful geyser data in Yellowstone
National Park.  It's the `faithful` dataset available in
the base R distribution.  Use `help(faithful)` or `?faithful`
to see its documentation,

2) Plot the waiting time to the next eruptions vs. the
duration of the eruption,

3) Fit a simple linear regression model to predict waiting
time from the eruption duration,

4) Overlay the fitted model on the plot with a red line,

#### Solution ####

Look at the documentation.
```{r}
?faithful
```

Plot the data.
```{r}
with(faithful, plot(eruptions, waiting))
```

Fit the simple linear regression model and overlay a regression line, make it thick with the `lwd`.
(line width) argument.
```{r}
slrModel <- lm(waiting ~ eruptions, data = faithful)
```

```{r}
plot(slrModel, col = "Red", lwd = 5)
```

Shut down the graph.
```{r}
dev.off()
```

#### Exercise 4.2: Trellis plots ####
1) Attach the quakes data that comes with R. Look at the
columns using `head()` or `str()`.

2) Use `xyplot` to plot lat vs. long for the location of the
earthquakes. Label the axes and the main graphs
using the following command to bin the depth variable into
9 bins:

```r
depthbin <- equal.count(quakes$depth, number = 9, overlap = 0)
```

Make sure you have loaded the lattice package via:
`library(lattice)`.

3) Now use `xyplot()` to plot lat,long for each depthbin of
the earthquake to create a set of graphs in a trellis
display.

#### Solution ####

Attach quakes for convenience in accessing its columns:
```{r}
attach(quakes)
```

Check out the data:
```{r}
head(quakes)
str(quakes)
```

Create the depth binning variable:
```{r}
library(lattice)
depth_bin <- equal.count(quakes$depth, number = 9, overlap = 0)
```

Produce the plot:
```{r}
xyplot(lat ~long | depth_bin,
main = "Earthquakes in the Pacific",
xlab = "longitude", ylab = "latitude")
```

Detach the quakes data frame:
```{r}
detach(quakes)
```


